---
title: "Hwk1 - Install/run R, RStudio, Rmarkdown"
subtitle: "Summer 2021"
author: "Your Name.. I like that, it's annonymous.. however, I feel the credit would be also annonymous.. =/.. this is a conundrum. (Sasha DC)"
date: "May 20^{th} 2021 "
output:
  word_document:
    reference_docx: WordStyle115.docx
---

```{r setup, include=F}
# set default parameters for knitting (output)
# "include = F" means this particular chunk will now show in output
# but by default, the rest of the chunks will display

knitr::opts_chunk$set(echo = T,  warning=F, results="hold")
```

#Part 1

## Read file

Read in the hwk 1 data csv file from working directory.
```{r read hwk1 file}
# Note that every chunk must have a different name
# read file
hwk1.dat <- read.csv('hwk1-115.csv')
View(hwk1.dat)
summary(hwk1.dat)
```

## Plot data

Be sure to replace *<Your Name>* with your actual name.
```{r plot hwk1.dat}
title=c("Sasha is the **bestest** Coug EVVVAAAAARRRRRRR!!")
plot(hwk1.dat, asp=1, col="#981e32", pch=15, cex=.2, main=title, bty="n")
```

There are many graphics tools and options in R. You will learn more through the semester.

Always put your answers, discussions and interpretations outside of the chunks. Only put comments that directly related to the code inside chunks.

#Part 2

## Bullets

Bullet the statements below by starting each line with an asterisk. **Bold** "Coug". Put your name in *italics*.

* *Sasha* is a **Coug**.
* But Dave, *the video insturctor guy* thinks the Coug stuff is a bit silly.

## Block quote

Put the following statements in a **block quote** by starting each line with ">".

> If at first you don't suceed, try, try again.

> Or you can email the instructor for help. :)


## Equations

You are all familiar with the equation for the area of a circle, 

$$A = \pi*r^{2}$$

### Equation block

You can also set an equation in a block. Use two dollar signs on front and back to put Einstein's mass-energy equivalent equation E=mc2 in an equation block.

$$E=MC^{2}$$

**Note**: Always put an extra line after chunks and before style changes like bullets and block quotes.
